---
title: "Rides to Safety: Scrutinizing Crime in Hyde Park and Campus Transits"
author: "Harsh Vardhan Pachisia"
date: "`r Sys.Date()`"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


"Five or six guys came up to them and pointed guns to their heads and demanded that they grab all their stuff. Phone, wallet, keys to their house, all taken. Physically, they're fine, but they're pretty shaken up, pretty rattled" [(CBS News, 2023).](https://www.cbsnews.com/chicago/news/2-armed-robberies-reported-near-university-of-chicago-campus/)

Such quotes are unfortunately pretty common to be heard around the University of Chicago's campus. The Hyde Park neighborhood is perceived to be [unsafe](https://www.insidehighered.com/news/2021/12/15/u-chicago-confronts-increase-violent-crime-around-campus) due to the prevalence of systemic criminal activities. With its unique position of being a 'college town' in the second city, the university has implemented various services aimed at enhancing student safety. It has "one of the largest private security forces in America" [(Vice, 2014)](https://www.vice.com/en/article/4w7p8b/why-does-a-campus-police-department-have-jurisdiction-over-65000-chicago-residents-1112) that patrol over 65,000 residents and UChicago Safe (an application targeted at safety). In particular, to promote outdoor safety, it has a variety of 'safety transportation' programs such as UGO shuttle service, initiated in the 1950s and expanded in 2017, and the 2021 UChicago Free Lyft Program, offering seven free rides per month within Hyde Park. 

Despite such extensive coverage of police, crime at Hyde Park seems to consistently increase, with three students being killed in 2021 and dozens having been robbed [(CBS News, 2023)](https://www.cbsnews.com/chicago/news/2-armed-robberies-reported-near-university-of-chicago-campus/). I'm motivated to analyze this problem as a graduate student at UChicago with first hand experiences during my time in Hyde Park. With that aim, this article and application examines the evolution of crime rates in Hyde Park and evaluates the effectiveness of these safety transportation options in reducing outdoor crime.

From an anecdotal standpoint, a lot of the UGO shuttles routes do not cater to a large majority of students with people feeling unsafe while walking around campus. In short, I hypothesize that the transit options offered are not effective in improving safety, despite that being their primary aim. Moreover, the impending [discontinuation](https://chicagomaroon.com/41051/news/university-officials-consider-replacing-lyft-program-with-new-via-service/#:~:text=The%20shift%20away%20from%20the,to%20discuss%20the%20potential%20change.) of the Lyft Program underscores the urgency to assess its contribution to campus safety.

In this article, I first investigate how the incidence of crime in Hyde Park has evolved over time. Second, I analyze whether the increase in safety transportation options by the university have improved safety by reducing the incidence of outdoor crime in Hyde Park? 

# Data Source and Cleaning

To answer my research questions, I use the Chicago crime data present at [Chicago data portal](https://data.cityofchicago.org/Public-Safety/Crimes-2001-to-Present/ijzp-q8t2/about_data). The entire Chicago crime dataset covers crimes taking place from 2001 to present day and covers all Chicago neighborhoods. Each row in the dataset is one reported crime, with a total dataset size of close to 8 million observations. Due to my research questions, I have limited the data to the Hyde Park area by taking a geographic bounding box of the Hyde Park area, and crimes between January 2014-December 2023.

To clean the data, I first create various date columns (time of day, month, year) using the date timestamp attached to every crime (using the lubridate package). Second, I remove any rows that do not contain either the primary type (type of crime column) or the time of day when the crime was committed and crimes that took place after Dec 31, 2023. Third, to make the analysis simpler, I only keep the columns that will be required for our analysis. This gives me cleaned data set of 20,732 crimes that took place in Hyde Park during these 10 years. 

```{r, echo=FALSE, warning=FALSE, message=FALSE}
#packages needed
library(ggplot2)
library(tidyverse)
library(lubridate)
library(dplyr)
library(tidyr)
library(gridExtra)
library(grid)
library(leaflet)
library(leaflet.extras)
library(shiny)
library(sf) 
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# getting crime data in the UChicago area (community areas: Hyde Park)
uchicago_crime<- read.csv('data/uchicago_crime.csv', header = TRUE)
```

```{r, echo= FALSE, warning=FALSE, message=FALSE}
# getting a date column and subsetting data to only include crimes till end-2023 (removing crimes in Jan 2024)
uchicago_crime$date_of_crime <- as.Date(uchicago_crime$Date, format = "%m/%d/%Y") 
uchicago_crime<- uchicago_crime %>% 
  subset(date_of_crime <= "2023-12-31" & date_of_crime >= '2014-01-01')

# adding columns for time of day, month, and year
uchicago_crime$time_of_day <- hour(strptime(uchicago_crime$Date, "%m/%d/%Y %I:%M:%S %p"))
uchicago_crime$month <- months(strptime(uchicago_crime$Date, "%m/%d/%Y %I:%M:%S %p"))
uchicago_crime$year <- year(strptime(uchicago_crime$Date, "%m/%d/%Y %I:%M:%S %p"))

# Store bounding box coordinates
hp_bb <- c(
  left = -87.608221,
  bottom = 41.783249,
  right = -87.577643,
  top = 41.803038
)

uchicago_crime <- uchicago_crime %>%
  filter(Longitude >= hp_bb["left"], Longitude <= hp_bb["right"],
         Latitude >= hp_bb["bottom"], Latitude <= hp_bb["top"])

# filtering to only Hyde Park as the UChicago area
uchicago_crime <- uchicago_crime %>% 
  rename(
    id = ID,
    date = Date,
    primary_type = Primary.Type,
    time_of_day = time_of_day,
    location = Location.Description,
    latitude = Latitude,
    longitude = Longitude
  ) %>% 
  filter(!is.na(time_of_day) & !is.na(primary_type)) %>%
  select(time_of_day, primary_type, location, id, date, month, year, latitude, longitude)
```

# Analysis

### What crimes happen in Hyde Park?

The types of crime that take place at the University of Chicago are first investigated to paint the overall picture of crime in Hyde Park. 

#### Crimes in Hyde Park

```{r, echo= FALSE, warning=FALSE, message=FALSE, fig.width=12, fig.height=8}

# getting the top 10 crimes 
top_crimes <- uchicago_crime %>%
  count(primary_type) %>%
  top_n(9, n) %>%
  pull(primary_type) 

# plugging the rest into an 'Other Crimes' variable
uchicago_crime <- uchicago_crime %>%
  mutate(primary_type = ifelse(primary_type %in% top_crimes, primary_type, 'OTHER CRIMES'))

# getting the frequency of each crime
crime_type_frequencies <- uchicago_crime %>%
  count(primary_type) %>%
  arrange(desc(n))

# to allow for better readability in the plot
crime_type_frequencies$primary_type <- str_wrap(crime_type_frequencies$primary_type, width = 30)

# plot
ggplot(crime_type_frequencies, aes(x = reorder(primary_type, n), y = n)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Frequency of Crime Types in Hyde Park (Jan 2014 - Dec 2023)",
       subtitle = "Top 10 crime types shown",
       caption = "Note: 'Other crimes' contains all crimes that had less than 350 occurrences. 'Other Offense' is a distinct category in the dataset.\n Source: City of Chicago Crime Data",
       x = "Type of Crime",
       y = "Frequency of Crime") +
  theme(plot.title = element_text(size = 13),
        plot.subtitle = element_text(size = 7),
        plot.caption = element_text(size = 6))
```

Theft is by far the most evident type of crime in Hyde Park with over 6,000 occurrences in a 10 year period. Theft is a crime that often takes place outdoors, as robbery and burglary (according to Chicago's Police definition), are indoor crimes. The frequency of other top crimes, battery, criminal damage, etc. are bunched together around 2000 occurrences. This chart gives an overall background as to which crimes typically occur in the Hyde Park area. 


#### Crime types over time

To better understand such crimes, their occurrences over time is analyzed in the interactive chart below. By default, the top 5 crimes in Hyde Park have been selected, but you can analyze whichever crime interests you by clicking the drop down. 

```{r, echo= FALSE, warning=FALSE, message=FALSE}
crime_counts <- readRDS("data/crime_counts.rds")

inputPanel(
  selectInput("crime_type", "Select Crime Types:",
            choices = unique(crime_counts$primary_type),
            selected = c("theft",'battery', 'criminal damage', 'deceptive practice', 'motor vehicle theft'),
            multiple = TRUE)
)
plotOutput("density")

output$density <- renderPlot({
  specificCrimes <- tolower(input$crime_type)

  d <- filter(crime_counts, primary_type %in% specificCrimes) %>% 
    mutate(label = if_else(year == max(year), as.character(primary_type), NA_character_))
  
  if (nrow(d) > 0) {
    ggplot(d, aes(x=year, y=count, color=primary_type)) +
      geom_line() +
      labs(title = "Crimes over years", x = "Year", y = "Number of Crimes", color = "Crime Type") +
      scale_color_viridis_d() +
      theme_bw()
  } else {
    ggplot() + 
      annotate("text", x = 1, y = 1, label = "No data available for selected crime types", size = 6, hjust = 0.5, vjust = 0.5)
  }
})
```

In the top 5 crime types, all of them dropped in 2020 due to the pandemic. Post pandemic, there is a concerning increase in theft and motor vechicle theft, with both reaching new peaks in 2023. Criminal damage, battery and deceptive practice have also come back to pre-pandemic levels.  


### When do crimes happen? 

Given the types of crime, it's critical to understand when such crimes happen to justify the introduction of safety transportation options. Moreover, due to the school year, the analysis is split between the summer months (June, July, August) and the academic year since there should be seasonality in crime. The following heat map charts out what time of day the major crimes take place in Hyde Park. Due to the difference in time periods, crime counts are normalized. Since theft is a major outlier, the charts are faceted to showcase this relationship with and without theft. 

```{r,echo= FALSE, warning=FALSE, message=FALSE, fig.width=12, fig.height=8}

# Calculate days in each period
days_in_year <- 365
days_in_summer <- 31 + 31 + 30 # June, July, August
days_in_school_year <- days_in_year - days_in_summer

# Creating two datasets: one including theft and one excluding theft
crime_heatmap_with_theft <- uchicago_crime %>% 
  mutate(summer = ifelse(month %in% c("June", "July", "August"), "Summer", "School Year")) %>%
  group_by(time_of_day, summer, primary_type) %>% 
  summarise(crime_count = n()) %>%
  mutate(crime_count_normalized = ifelse(summer == "Summer", 
                                         crime_count / days_in_summer,
                                         crime_count / days_in_school_year)) %>%
  select(-crime_count) %>% 
  top_n(15, crime_count_normalized) %>%
  spread(key = primary_type, value = crime_count_normalized, fill = 0)

    
crime_heatmap_without_theft <- uchicago_crime %>% 
  filter(primary_type != "THEFT") %>%
  mutate(summer = ifelse(month %in% c("June", "July", "August"), "Summer", "School Year")) %>%
  group_by(time_of_day, summer, primary_type) %>% 
  summarise(crime_count = n()) %>%
  mutate(crime_count_normalized = ifelse(summer == "Summer", 
                                         crime_count / days_in_summer,
                                         crime_count / days_in_school_year)) %>%
  select(-crime_count) %>% 
  top_n(15, crime_count_normalized) %>%
  spread(key = primary_type, value = crime_count_normalized, fill = 0)

# Converting the data to long format for ggplot
long_crime_data_with_theft <- crime_heatmap_with_theft %>% 
  gather(key = primary_type, value = crime_count_normalized, -c(time_of_day, summer))

long_crime_data_without_theft <- crime_heatmap_without_theft %>% 
  gather(key = primary_type, value = crime_count_normalized, -c(time_of_day, summer))

unique_values_with_theft <- unique(long_crime_data_with_theft$time_of_day)

time_levels <- as.character(0:23)

# Convert time_of_day to a factor
long_crime_data_with_theft$time_of_day <- factor(long_crime_data_with_theft$time_of_day, 
                                                 levels = time_levels, 
                                                 ordered = TRUE)

long_crime_data_without_theft$time_of_day <- factor(long_crime_data_without_theft$time_of_day, 
                                                    levels = time_levels, 
                                                    ordered = TRUE)

# Defining custom breaks and labels for the x-axis
custom_breaks <- time_levels[seq(1, 23, 5)]  # Every 5 hours starting from 0
custom_labels <- format(strptime(custom_breaks, format = "%H"), format = "%I %p")  # Convert to 12-hour AM/PM format

# Creating the heatmap for data including theft
plot_with_theft <- ggplot(long_crime_data_with_theft, 
                          aes(x = time_of_day, y = primary_type, fill = crime_count_normalized)) +
  geom_tile() +
  scale_fill_gradient(low = "floralwhite", high = "steelblue", name = "Normalized Crime Count") +
  scale_x_discrete(breaks = custom_breaks, labels = custom_labels) +
  facet_wrap(~summer, scales = "free") +
  theme_minimal() +
   theme(
    plot.title = element_text(size = 15, hjust = 0.5),
    plot.subtitle = element_text(size = 9, hjust = 0.5),
  ) +
  labs(title = "Crime Types vs Time of Day (with Theft)",
        subtitle = "Normalized counts, with Theft",
       x = "Hour of Day",
       y = "Crime Type")

# Creating the heatmap for data excluding theft
plot_without_theft <- ggplot(long_crime_data_without_theft, 
                             aes(x = time_of_day, y = primary_type, fill = crime_count_normalized)) +
  geom_tile() +
  scale_fill_gradient(low = "floralwhite", high = "steelblue", name = "Normalized Crime Count") +
  scale_x_discrete(breaks = custom_breaks, labels = custom_labels) +
  facet_wrap(~summer, scales = "free") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 15, hjust = 0.5),
    plot.subtitle = element_text(size = 9, hjust = 0.5),
  ) +
  labs(title = "Crime Types vs Time of Day (without Theft)",
       subtitle = "Normalized counts, without Theft",
       x = "Hour of Day",
       y = "Crime Type",
       fill = "Normalized Crime Count")

# Creating a blank grob (graphical object) for spacing
blank_grob <- grobTree(rectGrob(gp = gpar(col = NA)))

# Combine the plots with space in between
grid.arrange(
  plot_with_theft,
  blank_grob,  # Add blank grob for spacing
  plot_without_theft,
  ncol = 1,    # Arrange plots in one column
  heights = c(1, 0.05, 1) 
)
```

There are two major findings that stand out from these charts. First, it is interesting that after normalization, the pattern of crimes taking place in the summer and the school year remain roughly the same. While there is literature that states that crime does peak in the summer months across the country, the Hyde Park area is often referred to as a 'ghost town' during those months due to the lack of university students on campus. Hence, seeing such similar patterns goes against conventional wisdom.

Second, a lot of crime types such as theft, deceptive practice, and battery peak in the afternoon. After talking to a few people that have worked in the Chicago police, I found that sometimes a crime taking place at 2 am gets logged into the system the next afternoon due to bureaucratic paperwork. Hence, we should consider the peaks of crime provided by this data with a grain of salt. Finally, I see that other crimes such as robbery and criminal damage peak in the evening/nighttime, which should be expected.

### Where do crimes happen?

The geographical hot spots of crime in Hyde Park can be seen via the interactive map below.

```{r,echo= FALSE, warning=FALSE, message=FALSE}

# Load data
crime_lat_long <- readRDS("data/crime_lat_long.rds")
university_poi <- readRDS("data/university_poi.rds")
residence_poi <- readRDS("data/residence_poi.rds")
bus_routes <- readRDS("data/bus_routes.rds")
uchicago_shuttles <- readRDS("data/uchicago_shuttles.rds")

inputPanel(
  sliderInput("year", "Year", 
              min = 2014, max = 2023,
              value = 2023, step = 1, sep = ""),
   checkboxInput("showCampus", "Major Campus Buildings", value = TRUE),
  checkboxInput("showResidence", "Major Residential Buildings", value = TRUE)
)

leafletOutput("hydeParkMap1", width = "100%", height = "500px")


 # have to generate heatmaps here since cant do spatial stuff outside app.R, doesn't deploy
  years <- unique(crime_lat_long$year)
  crime_lat_long <- crime_lat_long %>% 
    # count of crime greater than or equal to 3
    filter(count >=3)
  

  # Pre-generate heatmaps for each year
  pre_generated_maps <- lapply(years, function(y) {
    data_year <- crime_lat_long[crime_lat_long$year == y, ]
    m <- leaflet(data_year) %>%
      addProviderTiles(providers$CartoDB.Positron) %>%
      addHeatmap(lng = ~longitude, lat = ~latitude, intensity = ~count, radius = 15, blur = 20, max = 0.05
                 )
    saveRDS(m, paste0("data/heatmap_", y, ".rds"))
  })

  output$hydeParkMap1 <- renderLeaflet({
    year_selected <- input$year
    heatmap_file <- paste0("data/heatmap_", year_selected, ".rds")
    
    map <- if (file.exists(heatmap_file)) {
      readRDS(heatmap_file)
    } else {
      leaflet() %>%
        addProviderTiles(providers$CartoDB.Positron) %>%
        setView(lng = -87.5903, lat = 41.7943, zoom = 14)
    }
    
    # Conditionally add the POI layer based on user input
    if (input$showCampus) {
      map <- map %>%
        addAwesomeMarkers(
          lng = university_poi$longitude, 
          lat = university_poi$latitude, 
          popup = university_poi$name,
          icon = awesomeIcons(
            icon = 'university', 
            iconColor = 'white', 
            library = 'fa', 
            markerColor = 'blue'
          )
        )
        
    }
    # Conditionally add the POI layer based on user input
    if (input$showResidence) {
      map <- map %>%
        addAwesomeMarkers(
          lng = residence_poi$longitude, 
          lat = residence_poi$latitude, 
          popup = residence_poi$name,
          icon = awesomeIcons(
            icon = 'home', 
            iconColor = 'white', 
            library = 'fa', 
            markerColor = 'blue'
          )
        )
    }
    color_palette <- colorNumeric(
      palette = c('#0000FF', '#00FFFF', '#00FF00', '#FFFF00', '#FFA500', '#FF0000'),
      domain = crime_lat_long$count
    )
    leafletProxy("hydeParkMap1", data = crime_lat_long) %>% 
      addLegend(position = "bottomright",
                title = "Number of Crimes",
                pal = color_palette,
                values = crime_lat_long$count,
                opacity = 1)
    
    map
}
)
```

Crime in East Hyde Park and E Hyde Park Blvd is rapidly increasing over the years, which should be a major point of concern since a lot of university students live in those areas. Moreover, the streets near UChicago Medicine are consistently a hot spot of crime, which needs to be addressed. Finally, 60th and 61st street contain a number of hot spots, which are extremely close to a number of prominent university buildings (MACS, Harris, Law School, Crown).

Crimes that take place indoors cannot be affected by the safety transit programs put in place by the University. In the following chart, the number of crimes that actually take place 'outdoors', in streets, alleys, sidewalks and parking lots are analyzed. It is concerning that nearly 50% of all Hyde Park crime is taking place outdoors, making it critical that safety transit programs are improved over time to tackle this issue. 

```{r,echo= FALSE, warning=FALSE, message=FALSE}
outside_locations <- c("PARKING LOT / GARAGE (NON RESIDENTIAL)","SIDEWALK","STREET")

total_crime_counts <- uchicago_crime %>%
  group_by(location)  %>%
  count(location) %>% 
  arrange(desc(n)) %>% 
  ungroup() %>% 
  top_n(10, n)

# Calculating the total number of crimes
total_crimes <- sum(total_crime_counts$n)

# Adding a new column for the percentage
total_crime_counts <- total_crime_counts %>%
  mutate(percentage = n / total_crimes * 100,
         label = paste(location, sprintf("%.1f%%", percentage),sep ="\n"))


total_crime_counts <- total_crime_counts %>%
  mutate(location = ifelse(location %in% c("APARTMENT", "RESIDENCE"), "APARTMENT/RESIDENCE", location)) %>%
  mutate(location = ifelse(!(location %in% outside_locations) & !(location %in% c("APARTMENT/RESIDENCE")), "OTHER LOCATIONS", location)) %>% 
  mutate(location = ifelse(location %in% c("PARKING LOT / GARAGE (NON RESIDENTIAL)"), "PARKING LOT", location)) %>% 
  group_by(location) %>%
  summarise(n = sum(n)) %>% 
  arrange(desc(n)) %>% 
  mutate(percentage = n / total_crimes * 100,
         label = paste(location, sprintf("%.1f%%", percentage),sep ="\n"))


outside_locations <- c("PARKING LOT","SIDEWALK","STREET")

# Adding a new column to indicate location
total_crime_counts$crime_location <- ifelse(total_crime_counts$location %in% 
                                       outside_locations, 
                                       "OUTDOOR", "INDOOR")

ggplot(total_crime_counts, aes(x = reorder(location, percentage), y = percentage, fill = crime_location)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("OUTDOOR" = "steelblue", "INDOOR" = "maroon4")) +
  coord_flip() +
  theme_minimal() +
  labs(title = "Nearly 50% of crime in Hyde Park occurs outdoors",
       subtitle = "Top 5 locations for crimes around UChicago (Jan 2014 - Dec 2023) shown",
       caption = "Other locations' contains all locations that had less than 1000 occurrences\n Source: City of Chicago Crime Data",
       x = "Location of Crime",
       y = "Percent of total crime") +
  theme(plot.title = element_text(size = 15),
        plot.subtitle = element_text(size = 7),
        plot.caption = element_text(size = 6),
        legend.position="bottom",
        legend.title = element_blank())
```


### What are the effects of safety transit programs?

The findings of the previous visualizations highlight the need for safety transit programs such as the university shuttles and the Lyft program. Their effect is visualized below to better understand their impact since their introduction, plotting both indoor and outdoor crime trends via line charts, to get a quasi causal inference. 

```{r,echo= FALSE, warning=FALSE, message=FALSE}
# counting outdoor and indoor crimes by year
outside_crimes_by_year <- uchicago_crime %>% 
  filter(location %in% outside_locations) %>% 
  select(primary_type,location, year, month, time_of_day) %>% 
  count(year)

inside_crimes_by_year <- uchicago_crime %>% 
  filter(!(location %in% outside_locations)) %>% 
  select(primary_type, location, year, month, time_of_day) %>% 
  count(year)

# Combining the data sets
crimes_by_year <- bind_rows(
  mutate(outside_crimes_by_year, crime_location = "Outdoor"),
  mutate(inside_crimes_by_year, crime_location = "Indoor")
)
#Plotting the combined data
ggplot(crimes_by_year, aes(x = year, y = n, color = crime_location)) +
  geom_line(data = crimes_by_year %>% filter(crime_location == "Outdoor"), size = 1) +
  geom_line(data = crimes_by_year %>% filter(crime_location == "Indoor"), size = 1, linetype = "dashed") + 
  scale_color_manual(values = c("Outdoor" = "steelblue", "Indoor" = "maroon4")) +
  scale_linetype_manual(values = c("Outdoor" = "solid", "Indoor" = "dashed")) +  # Define linetypes manually
  theme_minimal(base_size = 14) + 
  theme(
    plot.title = element_text(size = 13),
    axis.title = element_text(size = 10),
    plot.caption = element_text(size = 6),
    axis.text.x = element_text(angle = 0, hjust = 1),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    legend.position = "bottom",
    legend.title = element_text(size = 9) 
  ) + 
  labs(title = "Increased transport options have not reduced outdoor crime in Hyde Park",
       caption = "Source: City of Chicago Crime Data",
       x = "Years",
       y = "Frequency of Crimes",
       color = "Location of Crime") +
  geom_vline(xintercept = 2021, linetype = "dashed", color = "deeppink") +
  geom_vline(xintercept = 2017, linetype = "dashed", color = "maroon") +
  annotate("text", x = 2021, y = max(crimes_by_year$n), label = "Lyft Program starts", 
           hjust = 1.1, vjust = 2, color = "deeppink", angle = 0, size = 3, fontface = "italic") +
  annotate("text", x = 2017, y = max(crimes_by_year$n), label = "UGO Shuttles expand", 
           hjust = 1.1, vjust = 2, color = "maroon", angle = 0, size = 3, fontface = "italic")
```

We see a quasi difference-in-difference with both indoor and outdoor crimes having parallel trends before the expansion of the shuttles and the introduction of the Lyft program. This chart highlights the fact that neither the expansion of the UGO UChicago shuttles in 2017 nor the Lyft program introduction in 2021 has had any impact on outdoor crimes (crimes in streets, parking lots, sidewalks) in Hyde Park, showcasing that both programs are potentially failing to meet their objective of improving safety for students in Hyde Park. In fact, there is a rise in crime in Hyde Park after both programs are introduced, across both indoor and outdoor crimes.

It should be noted that it is possible that there are confounding variables that are not measured which could be impacting the findings. For example, after the introduction of such programs, the behavior of students could have changed, opting to stay out later under the assumption that they could use such transit options, ultimately leading to an overall increase in crime. 

Next, to get a granular understanding of this issue, the major shuttle and bus routes are overlaid on the crime hot spots to highlight any hot spots that are not covered by a pre-existing route. 

```{r,echo= FALSE, warning=FALSE, message=FALSE}
#load data
crime_lat_long_data <- readRDS("data/crime_lat_long.rds")
university_poi_data <- readRDS("data/university_poi.rds")
residence_poi_data <- readRDS("data/residence_poi.rds")
bus_routes_data <- readRDS("data/bus_routes.rds")
uchicago_shuttles_data <- readRDS("data/uchicago_shuttles.rds")

inputPanel(
  sliderInput("selectedYear", "Year", 
              min = 2014, max = 2023,
              value = 2023, step = 1, sep = ""),
  checkboxInput("showCTABusRoutes", "CTA Buses (171 and 172)", value = TRUE),
  checkboxInput("showUChicagoShuttleRoutes", "UChicago Shuttles", value = TRUE)
)

leafletOutput("hydeParkMap2", width = "100%", height = "500px")

# Generate heatmaps
years_list <- unique(crime_lat_long_data$year)
crime_lat_long_data_filtered <- crime_lat_long_data %>% 
    filter(count >= 3)
  
pre_generated_heatmaps <- lapply(years_list, function(year) {
    heatmap_data <- crime_lat_long_data_filtered[crime_lat_long_data_filtered$year == year, ]
    heatmap_map <- leaflet(heatmap_data) %>%
      addProviderTiles(providers$CartoDB.Positron) %>%
      addHeatmap(lng = ~longitude, lat = ~latitude, intensity = ~count, radius = 15, blur = 20, max = 0.05)
    saveRDS(heatmap_map, paste0("data/heatmap_", year, ".rds"))
})

output$hydeParkMap2 <- renderLeaflet({
    selected_year <- input$selectedYear
    heatmap_file_path <- paste0("data/heatmap_", selected_year, ".rds")
    
    if (file.exists(heatmap_file_path)) {
      leaflet_map <- readRDS(heatmap_file_path)
    } else {
      leaflet_map <- leaflet() %>%
        addProviderTiles(providers$CartoDB.Positron) %>%
        setView(lng = -87.5903, lat = 41.7943, zoom = 14)
    }

    # Adding bus routes and shuttles layers
    if (input$showCTABusRoutes && !is.null(bus_routes_data)) {
      leaflet_map <- leaflet_map %>%
        addPolylines(data = bus_routes_data,
                     color = 'darkred',
                     weight = 2,
                     opacity = 1)
    }

    if (input$showUChicagoShuttleRoutes && !is.null(uchicago_shuttles_data)) {
      convert_xyz_to_xy <- function(geometry) {
        # Extract coordinates and remove the Z dimension
        coords <- st_coordinates(geometry)
        coords <- coords[, 1:2] # Keep only X and Y
        
        # Create a new LineString or MultiLineString without the Z dimension
        if (st_geometry_type(geometry) == "MULTILINESTRING") {
          st_multilinestring(list(coords))
        } else {
          st_linestring(coords)
        }
      }
      uchicago_shuttles_data_xy <- st_sfc(lapply(st_geometry(uchicago_shuttles_data), convert_xyz_to_xy), crs = st_crs(uchicago_shuttles_data))
      uchicago_shuttles_new <- st_sf(st_drop_geometry(uchicago_shuttles_data), geometry = uchicago_shuttles_data_xy)
      
      leaflet_map <- leaflet_map %>%
        addPolylines(data = uchicago_shuttles_new,
                     color = 'darkblue',
                     weight = 2,
                     opacity = 1)
    }

    # Adding legends
    crime_color_palette <- colorNumeric(
      palette = c('#0000FF', '#00FFFF', '#00FF00', '#FFFF00', '#FFA500', '#FF0000'),
      domain = crime_lat_long_data_filtered$count
    )

    leaflet_map %>%
      addLegend(position = "bottomright",
                title = "Number of Crimes",
                pal = crime_color_palette,
                values = crime_lat_long_data_filtered$count,
                opacity = 1)
})
```

There are certain areas of Hyde Park that are not covered by the major transit options, including the crime hot spot near the university hospital, parts of East Hyde Park where crime has steadily been increasing over time, and on 60th-61st streets close to a variety of university buildings. These areas are systemic hot spots that need to be investigated with on-ground research. 

We can see that the programs are not meeting their goals. To improve the situation, I would recommend smarter investment in these programs to improve their overall effectiveness. For example, optimizing their routes based on where outdoor crimes are most prevalent could improve safety and cover hot spots that are currently not covered. These buses and shuttles are meant to provide last mile connectivity & safety, which they are failing to do. Further investments in other location specific programs such as increased University Police patrol and street stations should be also be looked at.

"I don't want to get robbed in the middle of the day or night. I came here to study, not to learn how to be safe" [(CBS News, 2023)](https://www.cbsnews.com/chicago/news/2-armed-robberies-reported-near-university-of-chicago-campus/) is a sentiment shared by a lot of students at the University of Chicago. While renowned for its academic excellence, efforts to improve safety around its Hyde Park campus need to be undertaken through programs that actually work. Ultimately, if the situation does not improve, every incoming student will simply continue being told: "Don't worry, that bit of weariness goes away. Everywhere you go, you have to look behind your back a little bit, that's just Hyde Park" [(ABC, 2023).](https://abc7chicago.com/university-of-chicago-robberies-crime-college-students-robbed-armed/13936369/) 


# Further investigation

For a complete investigation of each type of crime (temporal trends, peaks in a day, location of crime) and the geographical hot spots of crime, please explore this [interactive Shiny application](https://harshpachisia.shinyapps.io/uchicago-crimes-story/) to dig into and find possible other insights. For example, its shocking to see the number of sexual crimes that have taken place in hospitals in Hyde Park. If you have any questions or further insights on crime in Hyde Park, please feel free to [contact](https://www.harshpachisia.com/) me.
